---
title: "Variable Selection via Net Benefit with NBvarsel"
format:
  html:
    fig-width: 5.9
    fig-height: 5.9
    fig-dpi: 300
    toc: true
    number-sections: true
    code-fold: true
    theme: cosmo
    page-layout: full
    code-tools:
      source: false
      toggle: true
      caption: "Code"
    lightbox: true
execute:
  cache: true
  message: false
  warning: false
  error: false
  echo: true
embed-resources: true
vignette: >
  %\VignetteIndexEntry{Variable Selection via Net Benefit with NBvarsel}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| echo: false

library(NBvarsel)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(patchwork)
library(rms)
library(gt)
library(glmnet)
library(pROC)
library(ggsci)
```

# Introduction

Traditional variable selection methods (e.g., backward elimination, LASSO) optimise statistical performance metrics such as AUC or deviance. However, the best-performing model in statistical terms does not necessarily yield the best clinical utility when predictor costs (test harms) are accounted for. The `NBvarsel` package implements a variable selection framework based on cross-validated **Net Benefit**, the standard metric for clinical utility in decision curve analysis.

The core function `nb_varsel()` evaluates predictor subsets by their contribution to clinical decision-making, optionally adjusted for test harms. It supports:

- **Exhaustive** and **groupwise** (backward elimination) search strategies
- Predictor costs (per-variable or grouped)
- Restricted cubic splines for continuous predictors
- Interaction terms
- Permutation importance scores
- Parallel computation

# Simple illustration {#sec-simple}

::: {.callout-tip title="Reproducible section" collapse="true"}
This section is fully reproducible as the data generation is included in the code.
:::

Let us consider a hypothetical scenario with four predictors, two continuous ($X_1$, $X_3$) and two binary ($X_2$, $X_4$), used to predict a binary outcome $Y$. The true data-generating model is:

$$\text{logit}(Y) = -3 + 3X_1 + 2X_2 + 0.1X_3 - 0.05X_4$$ {#eq-dgm-simple}

In this model, $X_1$ and $X_2$ are strongly predictive, $X_3$ and $X_4$ are weakly predictive, and the outcome has a prevalence of approximately 64%. Each predictor is associated with a test harm: $X_1$ and $X_3$ (harm = 0.05), $X_2$ (harm = 0.025), and $X_4$ (harm = 0.00005).

## Data generation

```{r}
#| label: simple-data

set.seed(42)
n <- 1000

X1 <- rnorm(n, mean = 0, sd = 1)
X2 <- rbinom(n, size = 1, prob = 0.7)
X3 <- rnorm(n, mean = 0, sd = 1)
X4 <- rbinom(n, size = 1, prob = 0.5)

log_odds <- -3 + 3 * X1 + 2 * X2 + 0.1 * X3 - 0.05 * X4
prob <- plogis(log_odds)
Y <- rbinom(n, size = 1, prob = prob)

df <- data.frame(X1 = X1, X2 = X2, X3 = X3, X4 = X4, Y = Y)

harms <- c(X1 = 0.05, X2 = 0.025, X3 = 0.05, X4 = 0.00005)
```

## Backward elimination (baseline)

Using backward elimination based on Wald's statistic, as implemented in `fastbw()` of the `rms` package:

```{r}
#| label: simple-bw

model_full <- lrm(Y ~ X1 + X2 + X3 + X4, data = df)
bw_result <- fastbw(model_full)
sprintf("Backward elimination retains: %s", paste(bw_result$names.kept, collapse = ", "))
```

## Exhaustive net benefit variable selection

```{r}
#| label: simple-exhaustive

exhaustive <- nb_varsel(
  data = df,
  outcome_var = "Y",
  include_interactions = FALSE,
  costs = harms,
  cv_folds = 5,
  mode = "exhaustive",
  thresholds = 0.5,
  allow_parallel = FALSE,
  permutation = TRUE,
  splines = FALSE,
  n_knots = 3,
  verbose = FALSE
)
```

## Results

```{r}
#| label: tbl-train-subsets
#| tbl-cap: "Results for the simple illustration. Models are ranked by cost-adjusted Net Benefit."

tbl_train <- exhaustive$all_models

tbl_train |>
  select(Model, AUC, Brier, Total_Cost, Avg_Adj_Net_Benefit, Avg_Net_Benefit) |>
  arrange(-Avg_Adj_Net_Benefit) |>
  gt(id = "tb-train") |>
  fmt_number(
    columns = c("AUC", "Brier", "Avg_Adj_Net_Benefit", "Avg_Net_Benefit"),
    decimals = 3
  ) |>
  cols_label(
    Model = "Included predictors",
    AUC = "AUC",
    Brier = "Brier Score",
    Avg_Net_Benefit = "Avg. Net Benefit",
    Total_Cost = "Total Cost",
    Avg_Adj_Net_Benefit = "Avg. Adj. Net Benefit"
  )
```

The variable importance of each feature is shown in @fig-simple-illustration-1, illustrating the main contributors to utility. In @fig-simple-illustration-2, the all-subset plot shows the evolution of the average adjusted Net Benefit across all possible combinations. It is immediately apparent that including $X_1$ improves performance. This simple illustration demonstrates how the best subset of predictors in statistical terms does not necessarily translate to better cost-utility.

```{r}
#| label: fig-simple-illustration
#| fig-cap:
#|   - "Net benefit based variable importance for the simple illustration."
#|   - "All subset plot for the simple illustration."

VIF_plot(tbl_train) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

all_subset_plot(
  tbl_train,
  filter = 100,
  metric = "Avg_Adj_Net_Benefit",
  y_axis = "Adjusted Net Benefit"
)
```

# Case study — fabricated clinical data {#sec-case-study}

::: {.callout-tip title="Reproducible section" collapse="true"}
This section uses fabricated data designed to resemble a clinical prediction setting. All data generation code is included.
:::

We now demonstrate the methodology on a more realistic scenario. We simulate a dataset mimicking a heart failure readmission prediction setting with 10 predictors and a binary outcome (readmitted within 30 days vs. not). Predictors are grouped into three cost categories:

- **Clinical history** (no cost): patient age, prior admission, specialist centre, symptom severity
- **Cardiac imaging** (moderate cost): ejection fraction, wall thickness ratio, valve abnormalities, pericardial effusion, chamber dilation
- **Blood biomarker** (higher cost): NT-proBNP

## Data generation

```{r}
#| label: case-data

# Clean up simple illustration objects to avoid namespace conflicts
rm(list = setdiff(ls(), lsf.str()))

set.seed(8156)
n_train <- 3000
n_test <- 1500
n_total <- n_train + n_test

# Clinical history (free)
patient_age <- round(rnorm(n_total, mean = 68, sd = 12))
prior_admission <- rbinom(n_total, 1, 0.30)
specialist_centre <- rbinom(n_total, 1, 0.45)
symptom_severity <- rbinom(n_total, 1, 0.40)

# Cardiac imaging (moderate cost)
ejection_fraction <- pmin(75, pmax(10, rnorm(n_total, 42, 14)))
wall_thickness <- pmin(1, pmax(0.05, rbeta(n_total, 2, 4)))
valve_abnormalities <- rpois(n_total, lambda = 0.8)
pericardial_effusion <- rbinom(n_total, 1, 0.12)
chamber_dilation <- rbinom(n_total, 1, 0.22)

# Blood biomarker (higher cost)
nt_probnp <- round(pmax(20, rlnorm(n_total, meanlog = 6.5, sdlog = 1.4)))

# True model — strong signal for key predictors
lp <- -16.0 +
  0.07 * patient_age +
  3.0 * prior_admission +
  -0.12 * ejection_fraction +
  10.0 * wall_thickness +
  1.5 * valve_abnormalities +
  2.5 * pericardial_effusion +
  1.8 * chamber_dilation +
  0.9 * log2(nt_probnp) +
  # Weak / null predictors:
  0.03 * specialist_centre +
  0.02 * symptom_severity

readmitted <- rbinom(n_total, 1, plogis(lp))

clinical_df <- data.frame(
  readmitted, patient_age, nt_probnp, prior_admission,
  specialist_centre, ejection_fraction, valve_abnormalities,
  pericardial_effusion, chamber_dilation, symptom_severity,
  wall_thickness
)

training_data <- clinical_df[1:n_train, ]
test_data <- clinical_df[(n_train + 1):n_total, ]

sprintf("Training prevalence: %.1f%%", 100 * mean(training_data$readmitted))
sprintf("Test prevalence: %.1f%%", 100 * mean(test_data$readmitted))
```

## Cost structure

Costs are defined as grouped costs. The group cost is added once if *any* predictor from that group is included in the model.

```{r}
#| label: case-costs

prevalence <- mean(test_data$readmitted)

grouped_costs <- list(
  history = list(
    cost = 0,
    vars = c("patient_age", "prior_admission", "specialist_centre",
             "symptom_severity")
  ),
  imaging = list(
    cost = prevalence * 0.03,
    vars = c(
      "ejection_fraction", "wall_thickness",
      "valve_abnormalities", "pericardial_effusion", "chamber_dilation"
    )
  ),
  biomarker = list(
    cost = prevalence * 0.08,
    vars = c("nt_probnp")
  )
)
```

## Exhaustive net benefit variable selection

```{r}
#| label: case-exhaustive

vars <- names(training_data)

start_time <- Sys.time()
exhaustive_results <- nb_varsel(
  data = training_data,
  outcome_var = "readmitted",
  costs = grouped_costs,
  thresholds = seq(0.01, 0.2, by = 0.01),
  include_interactions = FALSE,
  cv_folds = 5,
  mode = "exhaustive",
  allow_parallel = TRUE,
  permutation = TRUE,
  splines = FALSE,
  verbose = FALSE
)
end_time <- Sys.time()

sprintf("Exhaustive search took: %s", format(end_time - start_time))
```

### Best model

```{r}
#| label: tbl-case-best
#| tbl-cap: "Best model from exhaustive search"

exhaustive_results$best_model_stats |>
  select(Model, AUC, Brier, Total_Cost, Avg_Adj_Net_Benefit, Avg_Net_Benefit) |>
  gt() |>
  fmt_number(
    columns = c("AUC", "Brier", "Total_Cost", "Avg_Adj_Net_Benefit", "Avg_Net_Benefit"),
    decimals = 3
  ) |>
  cols_label(
    Model = "Included predictors",
    AUC = "AUC",
    Brier = "Brier Score",
    Avg_Net_Benefit = "Avg. Net Benefit",
    Total_Cost = "Total Cost",
    Avg_Adj_Net_Benefit = "Avg. Adj. Net Benefit"
  )
```

### All models

```{r}
#| label: tbl-case-all
#| tbl-cap: "All models from exhaustive search"

all_models <- exhaustive_results$all_models
all_models$Rank_NB <- rank(-all_models$Avg_Net_Benefit, ties.method = "min")
all_models$Rank_adj_NB <- rank(-all_models$Avg_Adj_Net_Benefit, ties.method = "min")

tbl_data <- all_models |>
  select(
    Rank_NB, Rank_adj_NB, Model, AUC, Brier,
    Total_Cost, Avg_Adj_Net_Benefit, Avg_Net_Benefit
  )

tbl <- tbl_data |>
  gt() |>
  fmt_number(
    columns = c("AUC", "Brier", "Total_Cost", "Avg_Adj_Net_Benefit", "Avg_Net_Benefit"),
    decimals = 3
  ) |>
  cols_label(
    Rank_NB = "Rank (NB)",
    Rank_adj_NB = "Rank (Adj. NB)",
    Model = "Included predictors",
    AUC = "AUC",
    Brier = "Brier Score",
    Avg_Net_Benefit = "Avg. NB",
    Total_Cost = "Total Cost",
    Avg_Adj_Net_Benefit = "Avg. Adj. NB"
  )

if (knitr::is_html_output()) {
  tbl <- tbl |>
    opt_interactive(
      use_pagination = TRUE,
      page_size_default = 10,
      use_filters = TRUE,
      use_compact_mode = TRUE
    )
}
tbl
```

### Visualisation

```{r}
#| label: fig-case-vif
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-cap: "Variable importance plot (exhaustive search)"

VIF_plot(
  exhaustive_results$all_models,
  filter = round(nrow(exhaustive_results$all_models) * 0.1)
)
```

@fig-case-vif shows the variable importance based on the average Net Benefit contribution. Predictors are ranked from most to least important.

```{r}
#| label: fig-case-subset
#| fig-width: 6.4
#| fig-height: 8
#| fig-cap:
#|   - "All subset plot (Net Benefit)"
#|   - "All subset plot (Adjusted Net Benefit)"

all_subset_plot(
  exhaustive_results$all_models,
  filter = 7,
  size_dot = 1
)

all_subset_plot(
  exhaustive_results$all_models,
  filter = 7,
  size_dot = 1,
  metric = "Avg_Adj_Net_Benefit",
  y_axis = "Adjusted Net Benefit"
)
```

@fig-case-subset-1 and @fig-case-subset-2 show the all-subset plots. Each point represents a model. The heatmap below indicates which predictors are included. Note the difference in ranking when costs are accounted for.

## Groupwise net benefit variable selection

```{r}
#| label: case-groupwise

start_time <- Sys.time()
groupwise <- nb_varsel(
  data = training_data,
  outcome_var = "readmitted",
  costs = grouped_costs,
  thresholds = seq(0.01, 0.2, by = 0.01),
  include_interactions = FALSE,
  cv_folds = 5,
  mode = "groupwise",
  allow_parallel = TRUE,
  permutation = TRUE,
  splines = FALSE,
  group_size = 1,
  verbose = TRUE
)
end_time <- Sys.time()
sprintf("Groupwise search took: %s", format(end_time - start_time))
```

```{r}
#| label: tbl-case-groupwise
#| tbl-cap: "Groupwise selection: best model"

groupwise$best_model_stats |>
  select(Model, AUC, Brier, Total_Cost, Avg_Adj_Net_Benefit, Avg_Net_Benefit) |>
  gt() |>
  fmt_number(
    columns = c("AUC", "Brier", "Total_Cost", "Avg_Adj_Net_Benefit", "Avg_Net_Benefit"),
    decimals = 3
  ) |>
  cols_label(
    Model = "Included predictors",
    AUC = "AUC",
    Brier = "Brier Score",
    Avg_Net_Benefit = "Avg. NB",
    Total_Cost = "Total Cost",
    Avg_Adj_Net_Benefit = "Avg. Adj. NB"
  )
```

```{r}
#| label: fig-case-group-vif
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-cap: "Variable importance (groupwise selection)"

VIF_plot(groupwise$all_models)
```

## Comparison with other methods {#sec-comparison}

We compare the NB-based variable selection with backward elimination (Wald statistic) and LASSO (L1-penalised logistic regression).

```{r}
#| label: comparison-models

preds <- setdiff(vars, "readmitted")

# --- Backward elimination ---
fmla_full <- reformulate(termlabels = preds, response = "readmitted")
model_full <- lrm(fmla_full, data = training_data)
bw <- fastbw(model_full)
model_bw <- lrm(
  reformulate(termlabels = bw$names.kept, response = "readmitted"),
  data = training_data
)
test_data$bw_pred <- plogis(predict(model_bw, newdata = test_data))

# --- LASSO ---
x_train <- as.matrix(training_data[, preds])
y_train <- training_data$readmitted
x_test <- as.matrix(test_data[, preds])

set.seed(4738)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, family = "binomial")
lasso_coefs <- coef(cv_lasso, s = "lambda.1se")
lasso_kept <- rownames(lasso_coefs)[as.vector(lasso_coefs != 0)]
lasso_kept <- setdiff(lasso_kept, "(Intercept)")
test_data$lasso_pred <- as.vector(
  plogis(predict(cv_lasso, newx = x_test, s = "lambda.1se"))
)

# --- NB models ---
preds_best_nb <- str_trim(
  str_split(
    all_models |> arrange(-Avg_Net_Benefit) |> slice(1) |> pull(Model),
    ","
  )[[1]]
)
preds_best_adj <- str_trim(
  str_split(exhaustive_results$best_model_stats$Model, ",")[[1]]
)

model_nb <- lrm(
  reformulate(preds_best_nb, "readmitted"),
  data = training_data
)
model_adj_nb <- lrm(
  reformulate(preds_best_adj, "readmitted"),
  data = training_data
)
test_data$nb_pred <- plogis(predict(model_nb, newdata = test_data))
test_data$adj_nb_pred <- plogis(predict(model_adj_nb, newdata = test_data))
```

### Selected variables

```{r}
#| label: selected-vars

sprintf("NB model retains: %s", paste(preds_best_nb, collapse = ", "))
sprintf("Adjusted NB model retains: %s", paste(preds_best_adj, collapse = ", "))
sprintf("Backward elimination retains: %s", paste(bw$names.kept, collapse = ", "))
sprintf("LASSO retains: %s", paste(lasso_kept, collapse = ", "))
```

### Discrimination

```{r}
#| label: tbl-discrimination
#| tbl-cap: "AUC on test data for each variable selection method"

auc_results <- data.frame(
  Method = c("NB Model", "Adjusted NB Model", "Backward elimination", "LASSO"),
  AUC = c(
    fastAUC(test_data$nb_pred, test_data$readmitted),
    fastAUC(test_data$adj_nb_pred, test_data$readmitted),
    fastAUC(test_data$bw_pred, test_data$readmitted),
    fastAUC(test_data$lasso_pred, test_data$readmitted)
  ),
  n_predictors = c(
    length(preds_best_nb),
    length(preds_best_adj),
    length(bw$names.kept),
    length(lasso_kept)
  )
)

auc_results |>
  gt() |>
  fmt_number(columns = "AUC", decimals = 3) |>
  cols_label(n_predictors = "N predictors")
```

### Decision curve analysis

```{r}
#| label: fig-dca
#| fig-width: 6
#| fig-height: 4
#| fig-cap:
#|   - "Decision curve analysis comparing net benefit across methods."
#|   - "Decision curve analysis comparing harm-adjusted net benefit."

thresholds <- seq(0.01, 0.30, by = 0.005)
n_test_obs <- nrow(test_data)
y_test <- test_data$readmitted
prev <- mean(y_test)

calc_nb <- function(probs, y, thresholds) {
  n <- length(y)
  vapply(thresholds, function(pt) {
    pred_pos <- probs >= pt
    tp <- sum(y == 1 & pred_pos)
    fp <- sum(y == 0 & pred_pos)
    (tp / n) - (fp / n) * (pt / (1 - pt))
  }, numeric(1))
}

nb_all <- prev - (1 - prev) * (thresholds / (1 - thresholds))

dca_df <- bind_rows(
  data.frame(
    threshold = thresholds,
    net_benefit = calc_nb(test_data$nb_pred, y_test, thresholds),
    label = "NB Model"
  ),
  data.frame(
    threshold = thresholds,
    net_benefit = calc_nb(test_data$adj_nb_pred, y_test, thresholds),
    label = "Adjusted NB Model"
  ),
  data.frame(
    threshold = thresholds,
    net_benefit = calc_nb(test_data$bw_pred, y_test, thresholds),
    label = "Backward elimination"
  ),
  data.frame(
    threshold = thresholds,
    net_benefit = calc_nb(test_data$lasso_pred, y_test, thresholds),
    label = "LASSO"
  ),
  data.frame(
    threshold = thresholds,
    net_benefit = nb_all,
    label = "Treat all"
  ),
  data.frame(
    threshold = thresholds,
    net_benefit = 0,
    label = "Treat none"
  )
)

# Compute costs per method
cost_nb <- NBvarsel:::calculate_model_cost(preds_best_nb, grouped_costs)
cost_adj <- NBvarsel:::calculate_model_cost(preds_best_adj, grouped_costs)
cost_bw <- NBvarsel:::calculate_model_cost(bw$names.kept, grouped_costs)
cost_lasso <- NBvarsel:::calculate_model_cost(lasso_kept, grouped_costs)

dca_df <- dca_df |>
  mutate(
    cost = case_when(
      label == "NB Model" ~ cost_nb,
      label == "Adjusted NB Model" ~ cost_adj,
      label == "Backward elimination" ~ cost_bw,
      label == "LASSO" ~ cost_lasso,
      TRUE ~ 0
    ),
    adj_net_benefit = net_benefit - cost
  )

model_labels <- c(
  "NB Model", "Adjusted NB Model", "Backward elimination",
  "LASSO", "Treat all", "Treat none"
)
dca_df$label <- factor(dca_df$label, levels = model_labels)

# Raw NB
ggplot(dca_df, aes(x = threshold, y = net_benefit, color = label, linetype = label)) +
  geom_line(linewidth = 0.7) +
  theme_classic(base_size = 12) +
  labs(x = "Decision Threshold", y = "Net Benefit", color = "Strategy",
       linetype = "Strategy") +
  coord_cartesian(xlim = c(0, 0.30), ylim = c(-0.01, max(dca_df$net_benefit) * 1.05)) +
  scale_color_manual(values = c(
    "NB Model" = "#0072B5", "Adjusted NB Model" = "#BC3C29",
    "Backward elimination" = "#E18727", "LASSO" = "#20854E",
    "Treat all" = "black", "Treat none" = "black"
  )) +
  scale_linetype_manual(values = c(
    "NB Model" = "solid", "Adjusted NB Model" = "solid",
    "Backward elimination" = "solid", "LASSO" = "solid",
    "Treat all" = "dashed", "Treat none" = "dotted"
  ))

# Adjusted NB
ggplot(dca_df, aes(x = threshold, y = adj_net_benefit, color = label, linetype = label)) +
  geom_line(linewidth = 0.7) +
  theme_classic(base_size = 12) +
  labs(
    x = "Decision Threshold",
    y = "Harm-adjusted Net Benefit",
    color = "Strategy",
    linetype = "Strategy"
  ) +
  coord_cartesian(xlim = c(0, 0.30), ylim = c(-0.01, max(dca_df$adj_net_benefit) * 1.05)) +
  scale_color_manual(values = c(
    "NB Model" = "#0072B5", "Adjusted NB Model" = "#BC3C29",
    "Backward elimination" = "#E18727", "LASSO" = "#20854E",
    "Treat all" = "black", "Treat none" = "black"
  )) +
  scale_linetype_manual(values = c(
    "NB Model" = "solid", "Adjusted NB Model" = "solid",
    "Backward elimination" = "solid", "LASSO" = "solid",
    "Treat all" = "dashed", "Treat none" = "dotted"
  ))
```

@fig-dca-1 shows the standard decision curve analysis. @fig-dca-2 shows the same curves after subtracting each method's test costs, highlighting the advantage of cost-aware variable selection.

# Summary

This vignette demonstrated the `NBvarsel` workflow:

1. **`nb_varsel()`** identifies optimal predictor subsets by maximising cross-validated Net Benefit, optionally adjusted for test costs.
2. **`VIF_plot()`** visualises permutation-based variable importance.
3. **`all_subset_plot()`** provides a comprehensive view of model performance and predictor inclusion across all evaluated combinations.
4. **`fastAUC()`** offers a fast rank-based AUC calculation for external validation.

The key insight is that the statistically best-performing model may not be the most clinically useful when predictor costs are taken into account. Cost-aware variable selection via Net Benefit can lead to simpler, cheaper models that maintain or improve clinical utility.
